---
title: "Reproducible Research - Project 1"
author: "dscitbox"
date: "Sunday, October 19, 2014"
output: html_document
---
This file contains the solution to the first project in the Reproducible Research course

###Part I: number of steps per day
```{r load and prepare}
library(data.table)
library(plyr)
library(ggplot2)
dataset<-fread("activity.csv")
goodsteps<-subset(dataset, subset = !is.na(steps))
goodsteps$date<-as.Date(goodsteps$date)
# summarize the number of steps (exclude the records missing data)
steps<-ddply(goodsteps, .(date), summarise, sum(steps))
colnames(steps)<-c("Date", "Sum")
```

1. A histogram showing the distribution of the total number of steps per day
```{r, echo=FALSE}
m <- ggplot(steps, aes(x=Sum))+xlab("Total number of steps per day")+ylab("Frequency (number of days)")
m+geom_histogram(aes(fill = ..count..), binwidth=5000) + xlim(0, 25000)
```
2. The mean and median of the total of steps taken per day
```{r, echo=F}
meanTotal<-mean(steps$Sum)
medianTotal<-median(steps$Sum)
```
      the mean is `r format(round(meanTotal, 2), nsmall = 2)` and the median is `r format(round(medianTotal, 2), nsmall = 2)`

###Part II: average daily pattern

1. Time series plot of the 5-minute intervals and number of steps taken, averaged across all days
```{r daily patern}
intervalSteps<-(ddply(goodsteps, .(interval), summarize, steps=mean(steps)))
```
```{r, echo=FALSE}
ggplot(intervalSteps, aes(x=interval, y=steps)) + 
   geom_line()+
   xlab("Interval number")+
   ylab("Number of steps")
```

2. Finding the index of the 5-minute interval that contains the maximum number of steps

      the 5-minute interval that on average across all the days in the dataset, contains the maximum number of steps is `r intervalSteps[intervalSteps$steps==max(intervalSteps$steps), 1]`
      
###Part III: Inputting missing values
```{r missing values}
nmissingvalues<-nrow(dataset[is.na(steps)])
```
1. The number of missing values is `r nmissingvalues`
2. Filling the missing values with the average for that day
```{r filling na data}
# create a set with missing data from the original set
datasetwna<-dataset       
datasetwna$date<-as.Date(datasetwna$date)
# make the number of steps numeric to be able to hold the average
datasetwna$steps<-as.numeric(datasetwna$steps)
# join with the average number of steps/interval
x<-merge(x=datasetwna, y=intervalSteps, by="interval", all=T)
# create a new column to hold the fixed steps
x$steps<-x$steps.x
# replace na's with the average number of steps
x$steps[is.na(x$steps.x)]<-x$steps.y[is.na(x$steps.x)]
# repeat the calculations from the first part
fixedsteps<-ddply(x, .(date), summarise, sum(steps))
# fix the column names
colnames(fixedsteps)<-c("Date", "Sum")

```

Plot the number of steps with filled in data

```{r, echo=FALSE}
m <- ggplot(fixedsteps, aes(x=Sum))+xlab("Total number of steps per day")+ylab("Frequency (number of days)")
m+geom_histogram(aes(fill = ..count..), binwidth=5000) + xlim(0, 25000)
```

2. The mean and median of the total of steps taken per day
```{r, echo=FALSE}
meanTotal<-mean(fixedsteps$Sum)
medianTotal<-median(fixedsteps$Sum)

```
      the mean of the fixed set is `r format(round(meanTotal, 2), nsmall = 2)` and the median is `r format(round(medianTotal, 2), nsmall = 2)`
      
      the number of days with steps between 10 and 15k has increased to over 30, but the mean and the median have remained the same

### Part IV: Are there differences in activity patterns between weekdays and weekends
```{r}
# using the dataset with the number of steps fixed up
x$weekday<-ifelse(weekdays(x$date)!="Saturday" & weekdays(x$date)!="Sunday", "Weekday", "Weekend")
intervalSteps<-(ddply(x, .(interval, weekday), summarize, steps=mean(steps)))
```

```{r, echo=FALSE}
ggplot(intervalSteps, aes(x=interval, y=steps)) +
   facet_wrap(~weekday, nrow=2, ncol=1) +
   geom_line()+
   xlab("Interval number") +
   ylab("Number of steps")
```

The graphs show that the weekend and weekday profiles have a similar distribution, with slightly increased activity during the weekend; the activity is more concentrated at the beginning of the day. In the weekend the activity is more spread out during the day 